Computing in the Field of Rationals

### 2012-06-29 fields.tex
omake

\title{Negative and Fractional Types}

@@

\begin{verbatim}
revise the paragraph below to 
say something that's consistent with the 
presentation in the previous section; the
previous type system was the natural numbers;
we presented a natural language of terms
for that type system and a semantics in 
the rig category of finite
sets and bijections (two symmetric
monoidal structures and distributivity).
now we extend the type system to correspond
to the non-negative rational numbers;
we want to find the corresponding language
and semantic interpretation
\end{verbatim}

@@

%%%%%%%%%%%%%%%%%
\subsection{Non-determinism}

In order to interpret fractional types, we first lift the values and
combinators of {{Pi}} to the world of sets and relations. The lifting is
realized using a monad whose functor is powerset. Intuitively:
\begin{itemize}
\item instead of interpreting a type {{b}} as a finite set of values $\{ v_1,
  v_2, \ldots, v_{\modulusp{b}}\}$, we now interpret each type {{b}} as the
  powerset of such values; note that the type {{0}} now denotes
  $\{\emptyset\}$, i.e., it has one element which is the empty set;
  \item a {{Pi}}-value {{v}} of type {{b}} is lifted to a singleton set 
  $\{ v \}$;
\item each combinator {{c : b1 <-> b2}} is lifted to a \emph{symmetric}
  relation $R_c$ defined as follows:
\[\begin{array}{r@{\qquad}c@{\qquad}l}
\{ v_1, v_2, \ldots \} &R_c& \bigcup \{ c(v_i) \} 
\end{array}\]
For example, the combinator {{swap+ : 1+1 <-> 1+1}} is lifted to the relation $R$:
\[\begin{array}{r@{\qquad}c@{\qquad}l}
\emptyset &R& \emptyset \\
\{ \mathit{left}~() \} &R& \{ \mathit{right}~() \} \\
\{ \mathit{right}~() \} &R& \{ \mathit{left}~() \} \\
\{ \mathit{left}~(), \mathit{right}~() \} &R& \{ \mathit{left}~(), \mathit{right}~() \} 
\end{array}\]
\end{itemize}
Because every combinator corresponds to a bijection, the lifted relation is
symmetric.  Formally we have moved from the category of finite sets and
bijections to the category of finite sets and \emph{symmetric relations}.

Now that we have introduced the potential for non-determinism, it is possible
to introduce a type {{1/b}}, values {{1/v}}, and combinators {{eta*}} and
{{eps*}} that witness the following isomorphism:
%subcode{opsem} include main
% eta* &: 1 <-> (1/b) * b :& eps*
where {{b}} is not ``equivalent'' to {{0}} as detailed next.

Intuitively, we view a value {{1/v}} as a \emph{constraint} that is only
satisfied if it is matched with a value {{v'}} that is identical to
{{v}}. The type {{1/b}} is interpreted as the powerset of all the values
{{1/v}} with {{v}} being a {{Pi}}-value of type {{b}}. For example, the type
{{1/(1+1)}} denotes:
\[
\{ 
\emptyset, 
\{ 1/\mathit{left}~() \}, 
\{ 1/\mathit{right}~() \},
\{ 1/\mathit{left}~(), 1/\mathit{right}~() \} 
\}
\]
Of particular interest is that both the type {{0}} and the type {{1/0}}
denote $\{\emptyset\}$.

Given this interpretation of types and values, the realization of 
{{eta* : 1 <-> (1/b) * b}} is the relation $R$: 
\[\begin{array}{r@{\qquad}c@{\qquad}l}
\emptyset &R& \emptyset \\
\{ () \} &R& \{ (1/v,v) ~|~ v \mbox{~is~a~} \Pi\mbox{-value~of~type~} b \}
\end{array}\]
For example, at type $1+1$, the interpretation of {{eta*}} is the relation:
\[\begin{array}{r@{\qquad}c@{\qquad}l}
\emptyset &R& \emptyset \\
\{ () \} &R& \{ (1/\mathit{left}~(),\mathit{left}~()), 
                (1/\mathit{right}~(),\mathit{right}~()) \} 
\end{array}\]
At type $0$, the interpretation of {{eta*}} is:
\[\begin{array}{r@{\qquad}c@{\qquad}l}
\emptyset &R& \emptyset \\
\{ () \} &R& \emptyset 
\end{array}\]
which is not a symmetric relation. Hence the presence of {{eta*}} introduces
some ``non-reversible'' effect and moves us to the category of finite sets 
and (not necessarily symmetric) relations. The interpretation of 
{{eps* : (1/b) * b <-> 1}} is the relation:
\[\begin{array}{r@{\qquad}c@{\qquad}l}
\emptyset &R& \emptyset \\
\{ \{ (1/v,v) ~|~ v \mbox{~is~a~} \Pi\mbox{-value~of~type~} b \} &R& \{ () \} \\
\{ \{ (1/v,v') ~|~ v \mbox{~is~not~identical~to~} v' \} &R& emptyset 
\end{array}\]
For example, at type $0$, we have:
\[\begin{array}{r@{\qquad}c@{\qquad}l}
\emptyset &R& \emptyset 
\end{array}\]
At type $1+1$, we have:
\[\begin{array}{r@{\qquad}c@{\qquad}l}
\emptyset &R& \emptyset \\
\{ \{ (1/\mathit{left}~(),\mathit{left}~()) \} &R& \{ () \} \\
\{ \{ (1/\mathit{left}~(),\mathit{right}~()) \} &R& \emptyset \\
\{ \{ (1/\mathit{right}~(),\mathit{left}~()) \} &R& \emptyset \\
\{ \{ (1/\mathit{right}~(),\mathit{right}~()) \} &R& \{ () \} \\
\{ \{ (1/\mathit{left}~(),\mathit{left}~()), (1/\mathit{left}~(),\mathit{right}~()) \} &R&
  \emptyset \} \\
\{ \{ (1/\mathit{left}~(),\mathit{left}~()), (1/\mathit{right}~(),\mathit{left}~()) \} &R&
  \emptyset \} \\
\{ \{ (1/\mathit{left}~(),\mathit{left}~()), (1/\mathit{right}~(),\mathit{right}~()) \} &R&
  \{ () \} \} \\
\{ \{ (1/\mathit{left}~(),\mathit{right}~()), (1/\mathit{left}~(),\mathit{left}~()) \} &R&
  \emptyset \} \\
\{ \{ (1/\mathit{left}~(),\mathit{right}~()), (1/\mathit{right}~(),\mathit{left}~()) \} &R&
  \emptyset \} \\
\{ \{ (1/\mathit{left}~(),\mathit{right}~()), (1/\mathit{right}~(),\mathit{right}~()) \} &R&
  \emptyset \} \\
\{ \{ (1/\mathit{right}~(),\mathit{left}~()), (1/\mathit{left}~(),\mathit{left}~()) \} &R&
  \emptyset \} \\
\{ \{ (1/\mathit{right}~(),\mathit{left}~()), (1/\mathit{left}~(),\mathit{right}~()) \} &R&
  \emptyset \} \\
\{ \{ (1/\mathit{right}~(),\mathit{left}~()), (1/\mathit{right}~(),\mathit{right}~()) \} &R&
  \emptyset \} \\
\{ \{ (1/\mathit{right}~(),\mathit{right}~()), (1/\mathit{left}~(),\mathit{left}~()) \} &R&
  \{ () \} \} \\
\{ \{ (1/\mathit{right}~(),\mathit{right}~()), (1/\mathit{left}~(),\mathit{right}~()) \} &R&
  \emptyset \} \\
\{ \{ (1/\mathit{right}~(),\mathit{right}~()), (1/\mathit{right}~(),\mathit{left}~()) \} &R&
  \emptyset \} 
\end{array}\]

The category of finite sets and relations is a compact closed category. Each
object has a dual which is itself. The main coherence condition requires us
to check that:
\[\begin{array}{rcl}
b &\rightarrow& b * 1 \\
  &\rightarrow_{\eta^*}& b * (1/b * b) \\
  &\rightarrow& (b * 1/b) * b \\
  &\rightarrow_{\epsilon^*}& 1 * b \\
  &\rightarrow& b
\end{array}\]
For example, let's check one instance of this identity at the type $1+1$ and
the value {{left ()}}:
\[\begin{array}{rcl}
\{ \mathit{left}~() \} &\rightarrow& \{ (\mathit{left}~(), ()) \} \\
  &\rightarrow_{\eta^*}& \{ (\mathit{left}~(),(1/\mathit{left}~(),\mathit{left}~())), 
                            (\mathit{left}~(),(1/\mathit{right}~(),\mathit{right}~())) \} \\
  &\rightarrow& \{ ((\mathit{left}~(),1/\mathit{left}~()),\mathit{left}~())), 
                   ((\mathit{left}~(),1/\mathit{right}~()),\mathit{right}~())) \} \\
  &\rightarrow_{\epsilon^*}& \{ ((),\mathit{left}~()) \} \\
  &\rightarrow& \{ \mathit{left}~() \} 
\end{array}\]
At type $0$, we would get:
\[\begin{array}{rcl}
\emptyset &\rightarrow& \emptyset \\
  &\rightarrow_{\eta^*}& \emptyset \\
  &\rightarrow& \emptyset \\
  &\rightarrow_{\epsilon^*}& \emptyset \\
  &\rightarrow& \emptyset
\end{array}\]

%%%%%%%%%%%%%%%%
\subsection{Syntax}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Zero Totalized Fields}

Without the important side condition on division by zero, we would
have {{1<->1/0*0}} which implies {{1 <-> 0}} consequently collapsing
the entire structure.

While fields require the ``no division by zero'' constraint, alternate
algebraic structures such as meadows
\cite{Bergstra:2009:MES:1507774.1508000} have been studied which admit
division and negative without this constraint. Such structures are
called ``zero totalized fields'' and define multiplicative inverse of
{{0}}. The axiomatic presentation of meadows involve removing the
combinators corresponding the axiom {{0*b<->0}} (i.e. {{dist0}} and
{{factor0}}) and {{1<->1/b*b}} (i.e. {{eta*}} and {{eps*}}). Instead
meadows include axioms \emph{reflexivity}, {{refl}} and {{refr}}, and
\emph{restricted inverse}, {{ril}} and {{rir}}, which are:
%subcode{opsem} include main
% refl :& 1/1/b <-> b :& refr
% ril :& b*(b*1/b) <-> b :& rir

\noindent
Combinators {{refl}} and {{refr}} are primitive and can be defined as:
%subcode{opsem} include main
% refl 1/(1/v) &|-->& v
% refr v &|-->& 1/(1/v)

\noindent
Combinators {{ril}} and {{rir}} are defined as machine transitions
similar to {{eta*}}/{{eps*}}:
%subcode{opsem} include main
% <ril, (v,(v',1/v')), C> &|-->& {[ril, v, C]}
% <rir, v, C> &|-->& forall v':b.{[rir, (v, (v',1/v')), C]}

While not as commonly used as fields, zero totalization is useful not
just because we are spared from continuously checking the division
constraint, but also because many other structures such as the
category of vectors spaces do admit a dual to the 0 vector
space. However their categorization is not well developed and we
however lose the nice correspondence that {{eta*}}/{{eps*}} have with
compact closed structures.

@@

Charles Pinter in his book on Abstract Algebra talks about how
negatives were used by \emph{logisticians} of a 1000 years ago used
negative numbers as a ``useful fiction''.  The concept of negative
numbers were ``discovered''much later than fractionals. Like Joan Baez
says ``half an apple is easier to understand than a negative apple''.

Negative types should correspond to sets with negative arity. But what
does that mean? Various approaches have been proposed -- Loeb uses the
notion of multiplicity functions to express negative sized sets
[CITE]. These notions have been extended to Euler characteristic of
categories by Leinster, Joyal and Schanuel -- negative values are
explained sometimes as equalities relating otherwise independent
objects -- Baez tells a story about ``island'' and positive and
``bridges connecting islands'' as negatives (since the effectively
reduce the number of disconnected elements.)

We use the use the syntax {{-b}} for negative types and {{-v}} for
negative values and the axiom {{0 <-> (-b)+b}} as the basis for
negatives. In {{Pi}}/{{PiEE*}}, values flowed from left-to-right in
circuits. Negative types are negative in the sense that they flow in
the opposite direction.

@@

For convenience, we sometimes use the notations {{b1 - b2}} to
indicate the types {{b1 + (-b2)}}.  The types of the new constructs
are:
\vspace{-15pt}
\begin{multicols}{2}  
%subcode{opsem} include main
% eta+ &: 0 <-> (-b) + b :& eps+

@@

\begin{definition}[Size of a type]
The definition of size is extended to include negative types.
{{[^-b^] = -[^b^]}}
\end{definition}

@@

%subcode{opsem} include main
%! columnStyle = rclr
%! fwd = \triangleright
%! bck = \triangleleft
% {[eta+, right v, C, s]}^^^ &|-->& {[eta+, left (-v), C ]}
% {[eta+, left (-v), C, s]}^^^ &|-->& {[eta+, right v, C ]}
  \end{enumerate}

Consider the combinator {{c:(-1+1)+1<->1}} defined below. On providing
the value {{right ():(-1+1)+1}} we get the output {{():1}}. But on
supply input {{left (right ())}} evaluation stop on the reverse
interpreter with value {{left (left ()):(-1+1)+1}}.

By definition, since negative values flow in the opposite direction of
positive values, this is not surprising. Combinators can terminate in
the forward interpreter (with result of type {{b2}}) or in the reverse
interpreter (with result of type {{b1}}). To define an {{eval}} in a
consistent manner without worrying about state of the evaluator, we
transform every combinator of the form {{c:b1<->b2}} to one of the
form {{c:-b2+b1<->0}} by the construction {{action(c)=id(+)c(;)eps+}}.

\begin{definition}[{{eval(c, v)}}]
For {{c:b1<->b2}} and {{v:b1}} we define {{eval(c, v)=v'}} such that
\begin{enumerate}
\item {{v':-b2+b1}} and
\item {{<action(c), right v, []> |-->* <action(c),v',[]>^^^}}
\end{enumerate}
\end{definition}

\begin{definition}[{{c1=c2}}]
  For {{c1:b1<->b2}} and {{c2:b1<->b2}}, we define {{c1=c2}} if
  {{~forall v:b1.eval(c1,v) = eval(c2,v)}}.
\end{definition}

@@

%%%%%%%%%
\subsection{Categorical Structure}
% \label{sec:neg-constructions}

As before, we can show that {{PiEE+}} is a bimonoidal
category. Additionally it has a compact closed structure in the
{{(0,+)}} monoid.

\begin{proposition}[Compact Closed {{(0,+,-)}}]
  {{PiEE+}} is a compact closed category with over the symmetric
  monoid {{(0,+)}} with the dual of objects {{b}} give by {{-b}} with
  morphisms:

\begin{enumerate}
\item {{0 <-> -b + b}} given by {{eta+}}, called the `unit map'.
\item {{b + -b <-> 0}} given by {{swap+ (;) eps+}}, called the `counit map'.
\end{enumerate}

\begin{enumerate}
\item {{zeroi (;) swap+ (;) (id (+) eta+) (;) assocl+ (;) ((swap+ (;) eps+) (+) id) (;) zeroe = id : b <-> b}}
\item {{zeroi (;) (eta+ (+) id) (;) assocr+ (;) (id (+) (swap+ (;) eps+)) (;) swap+ (;) zeroe = id : -b <-> -b}}
\end{enumerate}

>>>>

\title{Computing in the Field of Rationals}

@@

@@

@@

Without the important side condition on division by zero, we would
have {{1<->1/0*0}} which implies {{1 <-> 0}} consequently collapsing
the entire structure. In this context, an important area for future
investigation is to consider alternatives to fields as a model for
fractionals and negatives -- particularly algebraic structures that
are ``zero totalized fields'' such as meadows
\cite{Bergstra:2009:MES:1507774.1508000}.

@@

@@

What is a negative type anyway?

@@

For convenience, we sometimes use the notations {{b1 - b2}} and
{{b1/b2}} to indicate the types {{b1 + (-b2)}} and {{b1 * (1/b2)}}
respectively.  The types of the new constructs are:
\vspace{-15pt}
\begin{multicols}{2}  
%subcode{opsem} include main
% eta+ &: 0 <-> (-b) + b :& eps+

@@

\begin{definition}[Size of a type]
The defintion of size is extended to include negative types.
{{[^-b^] = -[^b^]}}
\end{definition}

@@

%subcode{opsem} include main
%! columnStyle = rclr
%! fwd = \triangleright
%! bck = \triangleleft
% {[eta+, right v, C, s]}^^^ &|-->& {[eta+, left (-v), C ]}
% {[eta+, left (-v), C, s]}^^^ &|-->& {[eta+, right v, C ]}

  \end{enumerate}

@@

\begin{proposition}[Coherence]
